---
title: "County-level/Municipality-level Mortality Data"
author: "Kaustubh Wagh"
date: "2025-07-25"
format: html
execute:
  echo: true
  warning: false
editor: visual
---

```{r paths-config, include=FALSE}
# ── Project paths ─────────────────────────────────────────────────────────────
# All paths are resolved relative to the project root using {here}.
# Place a blank file named `.here` (or an `.Rproj` file) at the repo root so
# here::here() always anchors to the same location regardless of where this
# .qmd file lives within the project.

library(here)

# Raw data
path_inegi      <- here("data", "raw", "inegi_mexico")
path_conapo_yr  <- here("data", "raw", "conapo_mexico",
                        "Population projections by municipals by year 2018-23.csv")
path_conapo_age <- here("data", "raw", "conapo_mexico",
                        "Population projections by municipals by age sex group May 27, 2025.csv")
path_cdc_clean  <- here("data", "raw", "cdc_wonder_usa",
                        "Multiple Cause of Death 2018-2023 county level 2018-23 cleaned.csv")
path_cdc_age    <- here("data", "raw", "cdc_wonder_usa",
                        "Multiple Cause of Death 2018-2023 county level 2018-23 by age groups.csv")

# Shapefiles
path_shapefiles <- here("data", "external", "shapefiles")

# Processed outputs
path_proc       <- here("data", "processed")
```

## Mexico Border States

### Mortality Data

```{r setup1, warning=FALSE, echo==FALSE}

# Mexico Mortality Data Processing Function for Years 2018-2023
# Author: Processing INEGI mortality data
# Date: 2025


# Load and install required packages
install_and_load_packages <- function() {
  packages_needed <- c("sf", "ggplot2", "dplyr", "viridis", "leaflet", 
                      "DT", "scales", "foreign", "readr", "tools", 
                      "stringr", "tidyverse", "purrr", "tidyr")
  
  for(pkg in packages_needed) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      cat("Installing", pkg, "...\n")
      install.packages(pkg, dependencies = TRUE)
    }
    # Just use library without pos parameter
    library(pkg, character.only = TRUE)
  }
}

# Call the function
install_and_load_packages()

# Define constants
get_constants <- function() {
  # Border states (adjust as needed)
  border_states <- c("02", "05", "08", "19", "26", "28")  # Baja California,  Coahuila, Chihuahua,Nuevo León, Sonora, Tamaulipas
  
  # State names mapping (add all states as needed)
  state_names <- c(
    "01" = "Aguascalientes", "02" = "Baja California", "03" = "Baja California Sur",
    "04" = "Campeche", "05" = "Coahuila", "06" = "Colima", "07" = "Chiapas",
    "08" = "Chihuahua", "09" = "Ciudad de México", "10" = "Durango",
    "11" = "Guanajuato", "12" = "Guerrero", "13" = "Hidalgo", "14" = "Jalisco",
    "15" = "México", "16" = "Michoacán", "17" = "Morelos", "18" = "Nayarit",
    "19" = "Nuevo León", "20" = "Oaxaca", "21" = "Puebla", "22" = "Querétaro",
    "23" = "Quintana Roo", "24" = "San Luis Potosí", "25" = "Sinaloa",
    "26" = "Sonora", "27" = "Tabasco", "28" = "Tamaulipas", "29" = "Tlaxcala",
    "30" = "Veracruz", "31" = "Yucatán", "32" = "Zacatecas"
  )
  
  # Sex labels
  sex_labels <- c("1" = "Male", "2" = "Female", "9" = "Not specified")
  
  return(list(
    border_states = border_states,
    state_names = state_names,
    sex_labels = sex_labels
  ))
}

get_age_group_labels <- function() {
  c(
    "01" = "Less than 1 year",
    "02" = "1 year",
    "03" = "2 years",
    "04" = "3 years",
    "05" = "4 years",
    "06" = "5 to 9 years",
    "07" = "10 to 14 years",
    "08" = "15 to 19 years",
    "09" = "20 to 24 years",
    "10" = "25 to 29 years",
    "11" = "30 to 34 years",
    "12" = "35 to 39 years",
    "13" = "40 to 44 years",
    "14" = "45 to 49 years",
    "15" = "50 to 54 years",
    "16" = "55 to 59 years",
    "17" = "60 to 64 years",
    "18" = "65 to 69 years",
    "19" = "70 to 74 years",
    "20" = "75 to 79 years",
    "21" = "80 to 84 years",
    "22" = "85 to 89 years",
    "23" = "90 to 94 years",
    "24" = "95 to 99 years",
    "25" = "100 to 104 years",
    "26" = "105 to 109 years",
    "27" = "110 to 114 years",
    "28" = "115 to 119 years",
    "29" = "120 and over",
    "30" = "Not specified"
  )
}

# Main function to process mortality data for any year
process_mortality_data <- function(year, base_folder_path) {
  
  # Validate year
  if (!year %in% 2018:2023) {
    stop("Year must be between 2018 and 2023")
  }
  
  # Install and load packages
  install_and_load_packages()
  
  # Get constants
  constants <- get_constants()
  border_states <- constants$border_states
  state_names <- constants$state_names
  sex_labels <- constants$sex_labels
  age_group_labels <- get_age_group_labels()
  
  # Convert year to 2-digit format for file naming
  year_suffix <- sprintf("%02d", year %% 100)
  
  # Construct folder path for the specific year
  folder_path <- file.path(base_folder_path, paste0("defunciones_base_datos_", year, "_dbf"))
  
  cat("Processing data for year:", year, "\n")
  cat("Looking in folder:", folder_path, "\n")
  
  # Check if folder exists
  if (!dir.exists(folder_path)) {
    stop(paste("Folder does not exist:", folder_path))
  }
  
  # List all .dbf files in the folder
  dbf_files <- list.files(path = folder_path, pattern = "\\.dbf$", full.names = TRUE)
  
  if (length(dbf_files) == 0) {
    stop(paste("No .dbf files found in:", folder_path))
  }
  
  cat("Found", length(dbf_files), ".dbf files\n")
  
  # Read all dbf files into a named list of data frames
  dbf_data_list <- lapply(dbf_files, function(file) {
    cat("Reading:", basename(file), "\n")
    tryCatch({
      st_read(file, options = c("ENCODING=UTF-8"), quiet = TRUE)
    }, error = function(e) {
      cat("Error reading", basename(file), ":", e$message, "\n")
      return(NULL)
    })
  })
  
  # Remove NULL entries (failed reads)
  dbf_data_list <- dbf_data_list[!sapply(dbf_data_list, is.null)]
  
  # Name the list elements
  names(dbf_data_list) <- tools::file_path_sans_ext(basename(dbf_files[!sapply(dbf_data_list, is.null)]))
  
  # Extract datasets with year-specific naming
  defun_name <- paste0("DEFUN", year_suffix)
  catminde_name <- paste0("CATMINDE", year_suffix)  # Note: might just be "CATMINDE"
  listamex_name <- paste0("LISTAMEX", year_suffix)  # Note: might just be "LISTAMEX"
  catemlde_name <- paste0("CATEMLDE", year_suffix)
  capgpo_name <- paste0("CAPGPO", year_suffix)
  
  # Try different naming conventions
  possible_defun_names <- c(defun_name, paste0("DEFUN", year), "DEFUN")
  possible_catemlde_names <- c(catemlde_name, paste0("CATEMLDE", year), "CATEMLDE")
  
  DEFUN <- NULL
  for (name in possible_defun_names) {
    if (name %in% names(dbf_data_list)) {
      DEFUN <- dbf_data_list[[name]]
      cat("Using DEFUN dataset:", name, "\n")
      break
    }
  }
  
  CATEMLDE <- NULL
  for (name in possible_catemlde_names) {
    if (name %in% names(dbf_data_list)) {
      CATEMLDE <- dbf_data_list[[name]]
      cat("Using CATEMLDE dataset:", name, "\n")
      break
    }
  }
  
  if (is.null(DEFUN)) {
    stop("Could not find DEFUN dataset. Available datasets: ", paste(names(dbf_data_list), collapse = ", "))
  }
  
  if (is.null(CATEMLDE)) {
    warning("Could not find CATEMLDE dataset. Municipality names will not be available.")
  }
  
  # -------------------------------
  # Filter DEFUN for diabetes (20D, 43I) in border states
  # -------------------------------
  cat("Filtering data for diabetes cases in border states...\n")
  
  selected_data <- DEFUN %>%
    filter(
      LISTA_MEX %in% c("20D", "43I"),
      ENT_RESID %in% border_states
    )
  
  cat("Found", nrow(selected_data), "diabetes deaths in border states\n")
  
  # -------------------------------
  # Merge occupation descriptions (if available)
  # -------------------------------
  if ("OCUPACIONES" %in% names(dbf_data_list)) {
    selected_data <- selected_data %>%
      left_join(dbf_data_list[["OCUPACIONES"]], by = c("OCUPACION" = "CVE")) %>%
      rename(DESCRIP_OCUP = DESCRIP)
    cat("Added occupation descriptions\n")
  }
  
  # -------------------------------
  # Map codes to names
  # -------------------------------
  selected_data <- selected_data %>%
    mutate(
      STATE_NAME = state_names[as.character(ENT_RESID)],
      SEX_NAME = sex_labels[as.character(SEXO)],
       # keep EDAD_AGRU exactly as in microdata, but also add a readable label
    AGE_GROUP  = sprintf("%02d", as.integer(EDAD_AGRU)),
    AGE_GROUP_LABEL = dplyr::coalesce(get_age_group_labels()[AGE_GROUP], "Not specified"),
    AGE_GROUP_LABEL = factor(AGE_GROUP_LABEL, levels = get_age_group_labels()) 
    )
  
  # -------------------------------
  # Merge municipality names (if CATEMLDE is available)
  # -------------------------------
  if (!is.null(CATEMLDE)) {
    cat("Processing municipality data...\n")
    
    # Deduplicate CATEMLDE
    CATEMLDE_dedup <- CATEMLDE %>%
      distinct(CVE_ENT, CVE_MUN, .keep_all = TRUE)
    
    # Create municipality catalog
    municipality_catalog <- CATEMLDE_dedup %>%
      select(CVE_ENT, CVE_MUN, NOM_LOC) %>%
      distinct() %>% 
      filter(CVE_ENT %in% border_states) %>%
      filter(!CVE_MUN %in% c("000", "999"))
    
    # Create state lookup table
    state_lookup <- selected_data %>%
      select(ENT_RESID, STATE_NAME) %>%
      distinct()
    
    # Add state names to municipality_catalog
    municipality_catalog <- municipality_catalog %>%
      left_join(state_lookup, by = c("CVE_ENT" = "ENT_RESID"))
    
    # Merge with selected_data
    selected_data <- municipality_catalog %>%
      left_join(
        selected_data,
        by = c("STATE_NAME", "CVE_ENT" = "ENT_RESID", "CVE_MUN" = "MUN_RESID")
      ) %>%
      rename(
        ENT_RESID = CVE_ENT,
        MUN_RESID = CVE_MUN,
        MUNICIPALITY_NAME = NOM_LOC
      )
  } else {
    # If no municipality data, create placeholder
    selected_data <- selected_data %>%
      mutate(MUNICIPALITY_NAME = paste("Municipality", MUN_RESID))
  }
  
  # -------------------------------
  # Summarize data
  # -------------------------------
  cat("Creating summary tables...\n")
  
  # Total by municipality
  municipality_totals <- selected_data %>%
    group_by(ENT_RESID, MUN_RESID) %>%
    summarise(
      total_deaths = sum(LISTA_MEX %in% c("20D", "43I"), na.rm = TRUE),
      STATE_NAME = first(STATE_NAME),
      MUNICIPALITY_NAME = first(MUNICIPALITY_NAME),
      .groups = "drop"
    ) %>%
    select(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, total_deaths) %>%
    arrange(STATE_NAME, ENT_RESID, MUN_RESID, MUNICIPALITY_NAME, desc(total_deaths))
  
  # Total by key demographic breakdowns
  summary_table <- selected_data %>%
    group_by(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, SEXO, AGE_GROUP) %>%
    summarise(
      total_deaths = sum(LISTA_MEX %in% c("20D", "43I"), na.rm = TRUE),
      .groups = "drop" 
    ) %>%
    select(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, SEXO, AGE_GROUP, total_deaths) %>%
    arrange(STATE_NAME, MUNICIPALITY_NAME, desc(total_deaths))
  
  summary_table2 <- selected_data %>%
    group_by(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID) %>%
    summarise(
      total_deaths = sum(LISTA_MEX %in% c("20D", "43I"), na.rm = TRUE),
      .groups = "drop" 
    ) %>%
    select(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, total_deaths) %>%
    arrange(STATE_NAME, MUNICIPALITY_NAME, desc(total_deaths))
  # NEW: Totals by age group (state/mun)
    age_group_totals <- selected_data %>%
      group_by(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, AGE_GROUP, AGE_GROUP_LABEL) %>%
      summarise(
        total_deaths = sum(LISTA_MEX %in% c("20D", "43I"), na.rm = TRUE),
        .groups = "drop"
      ) %>%
      arrange(STATE_NAME, MUNICIPALITY_NAME, AGE_GROUP)
    
    # NEW: Totals by age group & sex (state/mun)
    age_sex_totals <- selected_data %>%
      group_by(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, AGE_GROUP, AGE_GROUP_LABEL, SEXO, SEX_NAME) %>%
      summarise(
        total_deaths = sum(LISTA_MEX %in% c("20D", "43I"), na.rm = TRUE),
        .groups = "drop"
      ) %>%
      arrange(STATE_NAME, MUNICIPALITY_NAME, AGE_GROUP, SEXO)
    
    # NEW: Totals by Municipality & Sex (without age breakdown)
    mun_sex_totals <- selected_data %>%
      mutate(YEAR = year) %>%
      group_by(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, SEXO, SEX_NAME, YEAR) %>%
      summarise(
        total_deaths = sum(LISTA_MEX %in% c("20D", "43I"), na.rm = TRUE),
        .groups = "drop"
      ) %>%
      arrange(STATE_NAME, MUNICIPALITY_NAME, YEAR, SEXO)
  
  # Return results
  results <- list(
    year = year,
    raw_data = selected_data,
    municipality_totals = municipality_totals,
    summary_table = summary_table,
    age_group_totals = age_group_totals, # NEW
    age_sex_totals = age_sex_totals,     # NEW
    mun_sex_totals = mun_sex_totals,     # NEW
    datasets_found = names(dbf_data_list)
  )
  
  cat("Processing completed for year", year, "\n")
  cat("Total diabetes deaths found:", sum(municipality_totals$total_deaths), "\n")
  
  return(results)
}

# Convenience function to process multiple years
process_multiple_years <- function(years, base_folder_path) {
  results_list <- list()
  
  for (year in years) {
    cat("\n", strrep("=", 50), "\n")
    cat("Processing year:", year, "\n")
    cat(strrep("=", 50), "\n")
    
    tryCatch({
      results_list[[as.character(year)]] <- process_mortality_data(year, base_folder_path)
    }, error = function(e) {
      cat("Error processing year", year, ":", e$message, "\n")
      results_list[[as.character(year)]] <- NULL
    })
  }
  
  return(results_list)
}


# Process single year
base_path <- path_inegi
# result_2023 <- process_mortality_data(2023, base_path)
# 
# result_2018 <- process_mortality_data(2018, base_path)
# result_2019 <- process_mortality_data(2019, base_path)

# Process multiple years
results_all <- process_multiple_years(2018:2023, base_path)

```

Age groups mgmt

```{r datamgmt, warning=FALSE, echo==FALSE}

# Access results
# municipality_totals_2023 <- result_2023$municipality_totals
# summary_table_2023 <- result_2023$summary_table
# age_group_totals_2023 <- result_2023$age_group_totals
# age_sex_totals_2023 <- result_2023$age_sex_totals
# 
# municipality_totals_2018 <- result_2018$municipality_totals
# 
# municipality_totals_2019 <- result_2019$municipality_totals


# Compare across years
for (year in names(results_all)) {
  if (!is.null(results_all[[year]])) {
    total_deaths <- sum(results_all[[year]]$municipality_totals$total_deaths)
    cat("Year", year, ": Total diabetes deaths =", total_deaths, "\n")
  }
}

.parse_year <- function(x) as.integer(stringr::str_extract(x, "\\d{4}"))

# LONG format
age_mun_year_long <- names(results_all) %>%
  map_dfr(function(nm) {
    res <- results_all[[nm]]
    if (is.null(res) || is.null(res$age_group_totals)) return(NULL)
    yr <- .parse_year(nm)
    res$age_group_totals %>%
      mutate(year = yr) %>%
      select(year, ENT_RESID, MUN_RESID, STATE_NAME, MUNICIPALITY_NAME,
             AGE_GROUP_LABEL, total_deaths)
  })

# age_mun_year_long %>%
#   count(ENT_RESID, MUN_RESID, STATE_NAME, MUNICIPALITY_NAME, AGE_GROUP_LABEL, year) %>%
#   filter(n > 1) %>%
#   arrange(desc(n))

name_lookup <- age_mun_year_long %>%
  group_by(ENT_RESID, MUN_RESID) %>%
  summarise(
    STATE_NAME        = first(STATE_NAME[!is.na(STATE_NAME) & STATE_NAME != ""]),
    MUNICIPALITY_NAME = first(MUNICIPALITY_NAME[!is.na(MUNICIPALITY_NAME) & MUNICIPALITY_NAME != ""]),
    .groups = "drop"
  )

# Join back to long dataset, overwriting names
age_mun_year_long_std <- age_mun_year_long %>%
  select(-STATE_NAME, -MUNICIPALITY_NAME) %>%
  left_join(name_lookup, by = c("ENT_RESID", "MUN_RESID"))

# age_mun_year_long_std %>% 
#    count(ENT_RESID, MUN_RESID, AGE_GROUP_LABEL) %>%
#   filter(n > 1)

# WIDE format
age_mun_year_wide <- age_mun_year_long_std %>%
  mutate(
    AGE_GROUP_LABEL = stringr::str_squish(AGE_GROUP_LABEL),
    MUNICIPALITY_NAME = stringr::str_squish(MUNICIPALITY_NAME),
    STATE_NAME = stringr::str_squish(STATE_NAME)
  ) %>%
  pivot_wider(
    names_from = year,
    names_prefix = "deaths_",
    values_from = total_deaths,
    values_fill = 0,
    values_fn = list(total_deaths = ~sum(.x, na.rm = TRUE))  # <- collapse duplicates
  ) %>%
  arrange(STATE_NAME, MUNICIPALITY_NAME, AGE_GROUP_LABEL) %>%
  mutate(total_2018_2023 = rowSums(across(starts_with("deaths_")), na.rm = TRUE))

# age_mun_year_wide %>% 
#    count(ENT_RESID, MUN_RESID, AGE_GROUP_LABEL) %>%
#   filter(n > 1)

# Process data for years 2018-2023
years <- 2018:2023

# Extract municipality totals for each year
municipality_data_list <- lapply(names(results_all), function(result_name) {
  year <- gsub("result_", "", result_name)
  data <- results_all[[result_name]]$municipality_totals
  data$year <- as.numeric(year)
  return(data)
})


# Combine all years into one dataset
combined_municipality_data <- bind_rows(municipality_data_list)

# Sum data across all years by municipality
# Assuming you want to sum numeric columns and keep municipality identifiers
municipality_totals_all_years <- combined_municipality_data %>%
  group_by(ENT_RESID, MUN_RESID) %>%
  summarise(
    STATE_NAME = first(STATE_NAME),
    MUNICIPALITY_NAME = first(MUNICIPALITY_NAME),
    # Keep deaths as separate columns by year
    deaths_2018 = sum(total_deaths[year == 2018], na.rm = TRUE),
    deaths_2019 = sum(total_deaths[year == 2019], na.rm = TRUE),
    deaths_2020 = sum(total_deaths[year == 2020], na.rm = TRUE),
    deaths_2021 = sum(total_deaths[year == 2021], na.rm = TRUE),
    deaths_2022 = sum(total_deaths[year == 2022], na.rm = TRUE),
    deaths_2023 = sum(total_deaths[year == 2023], na.rm = TRUE),
    # Sum other numeric columns
    across(where(is.numeric) & !matches("year"), sum, na.rm = TRUE),
    years_covered = paste(sort(unique(year)), collapse = ", "),
    n_years = n_distinct(year),
    .groups = "drop"
  )

# collapse >=85+ years inot single age group 
age_mun_year_wide_collapsed <- age_mun_year_long_std %>%
    # drop missing labels first
  filter(!is.na(AGE_GROUP_LABEL)) %>%
  mutate(
    AGE_GROUP_COLLAPSED = case_when(
      AGE_GROUP_LABEL %in% c("85 to 89 years",
                             "90 to 94 years",
                             "95 to 99 years",
                             "100 to 104 years",
                             "105 to 109 years",
                             "110 to 114 years",
                             "115 to 119 years",
                             "120 and over",
                             "Not specified") ~ "85+ years",
      TRUE ~ as.character(AGE_GROUP_LABEL)
    )
  ) %>%
  group_by(ENT_RESID, MUN_RESID, STATE_NAME, MUNICIPALITY_NAME,
           AGE_GROUP_COLLAPSED, year) %>%
  summarise(total_deaths = sum(total_deaths, na.rm = TRUE), .groups = "drop") %>%
  mutate(year_col = paste0("deaths_", year)) %>%
  select(ENT_RESID, MUN_RESID, STATE_NAME, MUNICIPALITY_NAME,
         AGE_GROUP_COLLAPSED, year_col, total_deaths) %>%
  tidyr::pivot_wider(names_from = year_col,
                     values_from = total_deaths,
                     values_fill = 0) %>%
  arrange(STATE_NAME, MUNICIPALITY_NAME, AGE_GROUP_COLLAPSED) %>%
  mutate(total_mor_2018_2023 = rowSums(across(starts_with("deaths_")), na.rm = TRUE))

# age_mun_year_wide_collapsed %>% 
#    count(ENT_RESID, MUN_RESID, AGE_GROUP_COLLAPSED) %>%
#   filter(n > 1)

age_mun_year_wide_collapsed

save(age_mun_year_wide_collapsed, file = "age_mun_year_mor_wide_mx.RData")

```

### Population Data

Overall

```{r setup2, warning=FALSE, echo==FALSE}

# Import the population CSV file
population_data <- read_csv(path_conapo_yr)

# Check the structure of the imported data
glimpse(population_data)

# Rename variables 
population_data <- population_data %>%
  rename(
    STATE_NAME = NOM_ENT,
    MUNICIPALITY_NAME = NOM_MUN,
    SEX = SEXO,
    Year = ANO,
    Population = POB_TOTAL
  )

# Display the renamed data structure
glimpse(population_data)

# Collapse sex and year data to get total population per municipality
municipality_pop_totals <- population_data %>%
  group_by(CLAVE, CLAVE_ENT, STATE_NAME, MUNICIPALITY_NAME) %>%
  summarise(
    # Population by individual years
    Population_2018 = sum(Population[Year == 2018], na.rm = TRUE),
    Population_2019 = sum(Population[Year == 2019], na.rm = TRUE),
    Population_2020 = sum(Population[Year == 2020], na.rm = TRUE),
    Population_2021 = sum(Population[Year == 2021], na.rm = TRUE),
    Population_2022 = sum(Population[Year == 2022], na.rm = TRUE),
    Population_2023 = sum(Population[Year == 2023], na.rm = TRUE),
    # Total population across all years
    Total_Population = sum(Population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # ENT_RESID: Add leading zero if single digit
    ENT_RESID = sprintf("%02d", CLAVE_ENT),
    # MUN_RESID: Extract last three digits of CLAVE
    MUN_RESID = sprintf("%03d", CLAVE %% 1000)
  )

# Display the first few rows of the collapsed data
head(municipality_pop_totals)

```

Age group mgmt

```{r popmgmt}

infile <- path_conapo_age

raw <- readr::read_csv(infile, show_col_types = FALSE)

dataset_tidy <- raw %>%
  # 1) Rename
  rename(
    STATE_NAME        = NOM_ENT,
    MUNICIPALITY_NAME = NOM_MUN,
    SEX               = SEXO,
    Year              = ANO,
    Population        = POB_TOTAL
  ) %>%
  # 2) Drop fecha if present
  select(-any_of("fecha")) %>%
  # 3) Standardize codes
  mutate(
    ENT_RESID = sprintf("%02d", CLAVE_ENT),
    MUN_RESID = sprintf("%03d", CLAVE %% 1000)
  ) %>%
  # 4) Pivot longer for age groups
  pivot_longer(
    cols = starts_with("POB_"),
    names_to  = "Age_Group_raw",
    values_to = "Population_AgeGroup"
  ) %>%
  mutate(
    Age_Groups = case_when(
      str_detect(Age_Group_raw, "_mm$") ~ {
        start_age <- str_extract(Age_Group_raw, "(?<=^POB_)\\d+")
        paste0(as.integer(start_age), "+ years")
      },
      str_detect(Age_Group_raw, "^POB_\\d+_\\d+$") ~ {
        mm <- str_match(Age_Group_raw, "^POB_(\\d+)_(\\d+)$")
        paste0(as.integer(mm[, 2]), " to ", as.integer(mm[, 3]), " years")
      },
      TRUE ~ Age_Group_raw
    )
  ) %>%
  # 5) Pivot wider: years as columns
  pivot_wider(
    id_cols = c(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, SEX, Age_Groups),
    names_from  = Year,
    values_from = Population_AgeGroup,
    names_prefix = "Pop_"
  ) %>%
  # 6) Collapse across sexes (sum M + F)
  group_by(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, Age_Groups) %>%
  summarise(across(starts_with("Pop_"), sum, na.rm = TRUE), .groups = "drop") %>%
  # 7) Keep only 2018–2023
  select(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, Age_Groups,
         Pop_2018:Pop_2023) %>% 
  mutate(Total_pop_2018_2023 = rowSums(across(Pop_2018:Pop_2023), na.rm = TRUE))

save(dataset_tidy, file = "age_mun_year_pop_wide_mx.RData")


```

### Crude Mortality Rate: Mexico

```{r CDR, warning=FALSE, echo==FALSE}

# Merge the population and Mortality datasets by ENT_RESID and MUN_RESID
Mexico_dm_mor_data <- municipality_totals_all_years %>%
  left_join(municipality_pop_totals %>% 
              select(-STATE_NAME, -MUNICIPALITY_NAME), 
            by = c("ENT_RESID", "MUN_RESID"))


# Assuming there's a deaths column in one of your datasets
Mexico_dm_mor_data <- Mexico_dm_mor_data %>%
  mutate(
    # Calculate crude death rate per 100,000
    Crude_Death_Rate = (total_deaths / Total_Population) * 100000,
    
    # Calculate 95% CI using Poisson distribution
    # Standard error for rate
    SE_rate = sqrt(total_deaths) / Total_Population * 100000,
    
    # 95% CI bounds
    CDR_Lower_CI = Crude_Death_Rate - 1.96 * SE_rate,
    CDR_Upper_CI = Crude_Death_Rate + 1.96 * SE_rate,
    
    # Alternative method using exact Poisson CI (more accurate for small counts)
    CDR_Lower_CI_Exact = qpois(0.025, total_deaths) / Total_Population * 100000,
    CDR_Upper_CI_Exact = qpois(0.975, total_deaths + 1) / Total_Population * 100000
  )

# Display summary of the calculated rates
summary(Mexico_dm_mor_data$Crude_Death_Rate)

# Display first few rows with the new calculated variables
head(Mexico_dm_mor_data %>% select(STATE_NAME, MUNICIPALITY_NAME, total_deaths, Total_Population, Crude_Death_Rate, CDR_Lower_CI, CDR_Upper_CI))

# Save as RData file (could save multiple objects) - Recommended for R
save(Mexico_dm_mor_data, file = "Mexico_dm_mor_data.RData")

```

### Mortality by Sex: Mexico

```{r combine_sex_stratified_years}

# Combine sex-stratified mortality data across all years (2018-2023)
combined_mun_sex_year <- bind_rows(
  lapply(names(results_all), function(year_name) {
    if (!is.null(results_all[[year_name]]) && 
        !is.null(results_all[[year_name]]$mun_sex_totals)) {
      results_all[[year_name]]$mun_sex_totals
    }
  })
) %>%
  arrange(STATE_NAME, MUNICIPALITY_NAME, YEAR, SEXO)

# View sample data
cat("\n=== Combined Sex-Stratified Mortality Data ===\n")
head(combined_mun_sex_year, 20)

# Summary statistics
cat("\nTotal records:", nrow(combined_mun_sex_year), "\n")
cat("Years covered:", min(combined_mun_sex_year$YEAR), "-", max(combined_mun_sex_year$YEAR), "\n")
cat("Unique states:", n_distinct(combined_mun_sex_year$STATE_NAME), "\n")
cat("Unique municipalities:", n_distinct(combined_mun_sex_year$MUNICIPALITY_NAME), "\n")
cat("Sex categories:", unique(combined_mun_sex_year$SEX_NAME), "\n")

# ========================================================================
# PREPARE POPULATION DATA BY SEX & YEAR
# ========================================================================

# Read the detailed population data by sex and year
# This is your CONAPO data with sex stratification
infile_pop_sex <- path_conapo_yr


population_data_by_sex <- read_csv(infile_pop_sex, show_col_types = FALSE) %>%
  rename(
    STATE_NAME = NOM_ENT,
    MUNICIPALITY_NAME = NOM_MUN,
    SEX = SEXO,
    YEAR = ANO,
    Population = POB_TOTAL,
    ENT_RESID = CLAVE_ENT
  ) %>%
  # Map sex codes to labels
  mutate(
    SEX_NAME = case_when(
      SEX == "HOMBRES" ~ "Male",
      SEX == "MUJERES" ~ "Female",
      TRUE ~ as.character(SEX)
    ),
    # Standardize codes
    ENT_RESID = sprintf("%02d", as.numeric(ENT_RESID)),
    MUN_RESID = sprintf("%03d", as.numeric(CLAVE) %% 1000)
  ) %>%
  select(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, YEAR, SEX, SEX_NAME, Population) %>%
  arrange( ENT_RESID, MUN_RESID,YEAR, SEX_NAME )

cat("\n=== Population Data by Sex & Year ===\n")
head(population_data_by_sex, 20)
cat("\nTotal rows:", nrow(population_data_by_sex), "\n")

# Create wide format: population by year for each municipality-sex combination
pop_wide <- population_data_by_sex %>%
  pivot_wider(
    id_cols = c(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID,MUN_RESID, SEX_NAME),
    names_from = YEAR,
    names_prefix = "Pop_",
    values_from = Population,
    values_fill = 0
  ) %>%
  mutate(
    Pop_2018_2023 = Pop_2018 + Pop_2019 + Pop_2020 + Pop_2021 + Pop_2022 + Pop_2023
  ) %>%
  arrange(STATE_NAME, MUNICIPALITY_NAME, SEX_NAME)

head(pop_wide, 10)

write.csv(pop_wide,
          "pop_wide_year_2018-2023.csv",
          row.names = FALSE)

# ========================================================================
# CALCULATE CRUDE MORTALITY RATES BY SEX BY YEAR
# ========================================================================

# ========== STEP 1: Prepare mortality data for rate calculation ==========

# Aggregate mortality by sex and year (across all years in combined data)
mortality_sex_year <- combined_mun_sex_year %>%
  rename(SEX_NAME = SEX_NAME) %>%
  select(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, YEAR, SEX_NAME, total_deaths) %>%
  arrange(ENT_RESID, MUN_RESID, YEAR, SEX_NAME)

# ========== STEP 2: Join mortality with population data ==========

# Create mortality-population dataset
mortality_pop_sex_year <- mortality_sex_year %>%
  left_join(
    population_data_by_sex %>% select(ENT_RESID, MUN_RESID, YEAR, SEX_NAME, Population),
    by = c("ENT_RESID", "MUN_RESID", "YEAR", "SEX_NAME"  )
  ) %>%
  # Handle missing population values
  mutate(
    Population = ifelse(is.na(Population), 0, Population),
    # Calculate crude mortality rate per 100,000
    CMR = ifelse(Population > 0, (total_deaths / Population) * 100000, NA)
  ) %>% 
  select(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID, YEAR, SEX_NAME, total_deaths, Population, CMR)

cat("\n=== CRUDE MORTALITY RATES BY MUNICIPALITY, SEX, AND YEAR ===\n")
head(mortality_pop_sex_year, 30)

# Summary statistics
cat("\n=== CMR Summary Statistics ===\n")
mortality_pop_sex_year %>%
  filter(!is.na(CMR) & CMR != 0) %>%
  summarise(
    Mean_CMR = mean(CMR, na.rm = TRUE),
    Median_CMR = median(CMR, na.rm = TRUE),
    Min_CMR = min(CMR, na.rm = TRUE),
    Max_CMR = max(CMR, na.rm = TRUE),
    SD_CMR = sd(CMR, na.rm = TRUE)
  ) %>%
  print()

# By sex
cat("\n=== CMR Summary by Sex ===\n")
mortality_pop_sex_year %>%
  filter(!is.na(CMR) & CMR != 0) %>%
  group_by(SEX_NAME) %>%
  summarise(
    Mean_CMR = mean(CMR, na.rm = TRUE),
    Median_CMR = median(CMR, na.rm = TRUE),
    Min_CMR = min(CMR, na.rm = TRUE),
    Max_CMR = max(CMR, na.rm = TRUE),
    Total_Deaths = sum(total_deaths, na.rm = TRUE),
    N_Municipality_Year = n(),
    .groups = "drop"
  ) %>%
  print()

# Save to CSV
write.csv(mortality_pop_sex_year,
          "diabetes_CMR_municipality_sex_year_2018-2023.csv",
          row.names = FALSE)

# Save as RData file (could save multiple objects) - Recommended for R
save(mortality_pop_sex_year, file = "Mexico_mor_pop_sex_year.RData")

cat("\nData saved to: diabetes_CMR_municipality_sex_year_2018-2023.csv\n")


# ========================================================================
# CALCULATE POOLED CMR (ALL YEARS COMBINED)
# ========================================================================

# ========== Aggregate mortality across all years (2018-2023) ==========

# Sum deaths by municipality and sex (all years combined)
mortality_pooled <- combined_mun_sex_year %>%
  mutate(SEX_NAME = if_else(is.na(SEX_NAME) | SEX_NAME == "" | SEX_NAME == "Not specified", "Male", SEX_NAME)) %>%
  group_by(ENT_RESID, MUN_RESID, SEX_NAME) %>%
  summarise(
    total_deaths_pooled = sum(total_deaths, na.rm = TRUE),
    n_years = n_distinct(YEAR),
    years_covered = paste(sort(unique(YEAR)), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(ENT_RESID, MUN_RESID, SEX_NAME)

# ========== Aggregate population across all years (2018-2023) ==========

# Average annual population for the period (sum population / 6 years)
population_pooled <- population_data_by_sex %>%
  group_by(ENT_RESID, MUN_RESID, SEX_NAME, STATE_NAME, MUNICIPALITY_NAME) %>%
  summarise(
    total_population_all_years = sum(Population, na.rm = TRUE),
    avg_annual_population = sum(Population, na.rm = TRUE) / 6,
    .groups = "drop"
  ) %>%
  arrange(ENT_RESID, MUN_RESID, SEX_NAME)

# ========== Join and calculate pooled CMR ==========

cmr_pooled_sex <- mortality_pooled %>%
  left_join(
    population_pooled,
    by = c("ENT_RESID", "MUN_RESID", "SEX_NAME")
  ) %>%
  mutate(
    # Pooled CMR per 100,000
    # Using average annual population as denominator
    CMR_pooled = ifelse(avg_annual_population > 0, 
                        (total_deaths_pooled / total_population_all_years) * 100000, 
                        NA)
  ) 
  

cat("\n=== POOLED CRUDE MORTALITY RATES (2018-2023 Combined) ===\n")
head(cmr_pooled_sex, 30)

# Summary statistics for pooled CMR
cat("\n=== Pooled CMR Summary Statistics (All Years 2018-2023) ===\n")
cmr_pooled_sex %>%
  filter(!is.na(CMR_pooled) & CMR_pooled != 0) %>%
  summarise(
    Mean_CMR = mean(CMR_pooled, na.rm = TRUE),
    Median_CMR = median(CMR_pooled, na.rm = TRUE),
    Min_CMR = min(CMR_pooled, na.rm = TRUE),
    Max_CMR = max(CMR_pooled, na.rm = TRUE),
    SD_CMR = sd(CMR_pooled, na.rm = TRUE)
  ) %>%
  print()

# By sex
cat("\n=== Pooled CMR by Sex (2018-2023) ===\n")
cmr_pooled_sex %>%
  filter(!is.na(CMR_pooled) & CMR_pooled != 0) %>%
  group_by(SEX_NAME) %>%
  summarise(
    Mean_CMR = mean(CMR_pooled, na.rm = TRUE),
    Median_CMR = median(CMR_pooled, na.rm = TRUE),
    Min_CMR = min(CMR_pooled, na.rm = TRUE),
    Max_CMR = max(CMR_pooled, na.rm = TRUE),
    Total_Deaths = sum(total_deaths_pooled, na.rm = TRUE),
    Avg_Population = sum(avg_annual_population, na.rm = TRUE),
    N_Municipalities = n(),
    .groups = "drop"
  ) %>%
  print()

# Save to CSV
write.csv(cmr_pooled_sex,
          "diabetes_CMR_pooled_municipality_sex_2018-2023.csv",
          row.names = FALSE)

cat("\nData saved to: diabetes_CMR_pooled_municipality_sex_2018-2023.csv\n")


```

### ASDMR: Mexico

```{r ASDMR}

# Age-standardized Diabetes Mortality Rate: Mexico
# 1. Standardize the age group column name
age_mun_year_wide_collapsed <- age_mun_year_wide_collapsed %>%
  rename(Age_Groups = AGE_GROUP_COLLAPSED)

county_count_by_state_mx <- age_mun_year_wide_collapsed %>%
  distinct(ENT_RESID, MUN_RESID) %>%   # make sure no duplicates
  count(ENT_RESID, name = "n_counties") %>%
  arrange(desc(n_counties))
county_count_by_state_mx

# Merge the population and Mortality datasets by ENT_RESID and MUN_RESID
Mexico_asdmr_data <- age_mun_year_wide_collapsed %>%
  left_join(dataset_tidy %>% 
              select(-STATE_NAME, -MUNICIPALITY_NAME), 
            by = c("ENT_RESID", "MUN_RESID","Age_Groups"))

 # age_mun_year_wide_collapsed %>%
 #  count(ENT_RESID, MUN_RESID, Age_Groups) %>%
 #  filter(n > 1)


# WHO World Standard Population (2000–2025)

who_std <- tibble(
  Age_Groups = c(
    "0 to 4 years", "5 to 9 years", "10 to 14 years", "15 to 19 years",
                 "20 to 24 years", "25 to 29 years", "30 to 34 years", "35 to 39 years",
                 "40 to 44 years", "45 to 49 years", "50 to 54 years", "55 to 59 years",
                 "60 to 64 years", "65 to 69 years", "70 to 74 years", "75 to 79 years",
                 "80 to 84 years", "85+ years"),
  Std_Weight = c(0.0886, 0.0869,0.0860,0.0847,0.0822,0.0793,0.0761,0.0715,0.0659,
                 0.0604,0.0537,0.0455,0.0372,0.0296,0.0221,0.0152,0.0091,0.0064))

# Merge WHO standard into merged dataset
merged_std <- Mexico_asdmr_data %>%
  left_join(who_std, by = "Age_Groups")

# Example: use Total_2018_2023 population and total deaths for that period
# You must have a "Deaths" column in merged_dataset
merged_std <- merged_std %>%
  mutate(
    rate = (total_mor_2018_2023 / Total_pop_2018_2023) * 100000
  )

# Calculate ASMR
asmr_county_mx <- merged_std %>%
  group_by(STATE_NAME, MUNICIPALITY_NAME, ENT_RESID, MUN_RESID) %>%
  summarise(
    ASMR = sum(rate * Std_Weight, na.rm = TRUE) / sum(Std_Weight[!is.na(rate) & !is.na(Std_Weight)], na.rm = TRUE),
    deaths_total= sum(total_mor_2018_2023, na.rm = TRUE),
    .groups = "drop"
  )


asmr_county_mx<- asmr_county_mx %>%
  bind_rows(
    tibble(
      STATE_NAME = "Sonora",
      MUNICIPALITY_NAME = "Onavas",
      ENT_RESID = "26",
      MUN_RESID = "044",      
      ASMR = 0,
      deaths_total = 0
    )
  )

# asmr_county_mx %>%
#   count(ENT_RESID, MUN_RESID) %>%
#   filter(n > 1)

head(asmr_county_mx)

save(asmr_county_mx, file = "county_asdmr_mx.RData")




```

## USA Border States

### Crude Mortality Rate: USA

```{r CDR2, warning=FALSE, echo==FALSE}

# Import the CSV file
USA_dm_mor_data <- read_csv(path_cdc_clean)

# Display basic information about the dataset
glimpse(USA_dm_mor_data)

# Reusable county-name cleaner
clean_county_name <- function(x) {
  x %>%
    # remove standard "Type, ST" suffices
    str_remove("\\s+(County|Parish|Borough|Census Area|Municipality|City and Borough),\\s*[A-Z]{2}$") %>%
    # also handle when no state suffix is present
    str_remove("\\s+(County|Parish|Borough|Census Area|Municipality|City and Borough)$") %>%
    # trim
    str_trim()
}


# Reusable code formatter → adds STATEFP, COUNTYFP, GEOID, COUNTRY
add_geo_codes <- function(df, state_col = STATEFP, county_col = COUNTYFP) {
  df %>%
    mutate(
      STATEFP  = sprintf("%02d", as.integer({{ state_col }})),
      COUNTYFP = sprintf("%03d", as.integer({{ county_col }}) %% 1000),
      GEOID    = paste0(STATEFP, COUNTYFP),
      COUNTRY  = "USA"
    )
}

## Rename columns and clean COUNTY_NAME
USA_dm_mor_data <- USA_dm_mor_data %>%
  rename(
    STATE_NAME = State,
    COUNTY_NAME = County,
    STATEFP = `State Code`,
    COUNTYFP = `County Code`
  ) %>%
  mutate(
    COUNTY_NAME = clean_county_name(COUNTY_NAME)
  ) %>%
  add_geo_codes(state_col = STATEFP, county_col = COUNTYFP)

# Quick checks
USA_dm_mor_data %>%
  select(STATE_NAME, COUNTY_NAME, STATEFP, COUNTYFP, GEOID, COUNTRY) %>%
  head(10)

str(USA_dm_mor_data)

USA_dm_mor_data %>%
  select(STATE_NAME, COUNTY_NAME, STATEFP, COUNTYFP, GEOID, COUNTRY, Deaths, Population) %>%
  head(15)

# Clean COUNTY_NAME: remove "County, XX" suffix (e.g., "Apache County, AZ" -> "Apache")
USA_dm_mor_data <- USA_dm_mor_data %>%
  mutate(
    COUNTY_NAME = str_remove(COUNTY_NAME, " County, [A-Z]{2}$")
  )

# Check the renamed and cleaned columns
USA_dm_mor_data %>%
  select(STATE_NAME, COUNTY_NAME, STATEFP, COUNTYFP) %>%
  head(10)

##Format STATEFP and COUNTYFP
  # Format STATEFP: make single digits double digits (e.g., 4 -> 04)
USA_dm_mor_data <- USA_dm_mor_data %>%
  mutate(
    STATEFP = sprintf("%02d", as.numeric(STATEFP))
  )

# Format COUNTYFP: keep only last three digits (e.g., 4001 -> 001)
USA_dm_mor_data <- USA_dm_mor_data %>%
  mutate(
    COUNTYFP = sprintf("%03d", as.numeric(COUNTYFP) %% 1000)
  )

# Check the formatting
USA_dm_mor_data %>%
  select(STATE_NAME, COUNTY_NAME, STATEFP, COUNTYFP) %>%
  head(10)

## 4. Create GEOID and COUNTRY variables

# Create GEOID variable (combination of STATEFP and COUNTYFP)
# This will be 5 digits (2 for state + 3 for county)
USA_dm_mor_data <- USA_dm_mor_data %>%
  mutate(
    GEOID = paste0(STATEFP, COUNTYFP),
    COUNTRY = "USA"
  )

# Display the final dataset structure
USA_dm_mor_data %>%
  select(STATE_NAME, COUNTY_NAME, STATEFP, COUNTYFP, GEOID, COUNTRY) %>%
  head(10)

# Show column names and types
str(USA_dm_mor_data)

# Display first few rows of key columns
USA_dm_mor_data %>%
  select(STATE_NAME, COUNTY_NAME, STATEFP, COUNTYFP, GEOID, COUNTRY, Deaths, Population) %>%
  head(15)

# Save as RData file (could save multiple objects) - Recommended for R
save(USA_dm_mor_data, file = "USA_dm_mor_data.RData")


```

### ASDMR: USA

```{r setup, warning=FALSE, include=FALSE}

# Age-standardized Diabetes Mortality Rate: USA
# ============================================================
# County-level ASMR (WHO 2000–2025) for your CDC file
# ============================================================

library(tidyverse)
library(readr)

# Reusable county-name cleaner
clean_county_name <- function(x) {
  x %>%
    # remove standard "Type, ST" suffices
    str_remove("\\s+(County|Parish|Borough|Census Area|Municipality|City and Borough),\\s*[A-Z]{2}$") %>%
    # also handle when no state suffix is present
    str_remove("\\s+(County|Parish|Borough|Census Area|Municipality|City and Borough)$") %>%
    # trim
    str_trim()
}

# Reusable code formatter → adds STATEFP, COUNTYFP, GEOID, COUNTRY
add_geo_codes <- function(df, state_col = STATEFP, county_col = COUNTYFP) {
  df %>%
    mutate(
      STATEFP  = sprintf("%02d", as.integer({{ state_col }})),
      COUNTYFP = sprintf("%03d", as.integer({{ county_col }}) %% 1000),
      GEOID    = paste0(STATEFP, COUNTYFP),
      COUNTRY  = "USA"
    )
}

# ---- 1) Load data
path <- path_cdc_age

df_raw <- read_csv(path, show_col_types = FALSE)

df <- df_raw %>%
  mutate(
    Deaths = suppressWarnings(as.numeric(Deaths)),
    Population = suppressWarnings(as.numeric(Population))
  ) %>%
  filter(!is.na(Population), Population > 0, !is.na(Deaths)) %>%
  mutate(
    `Five-Year Age Groups` = ifelse(`Five-Year Age Groups` %in% c("< 1 year","1-4 years"),
                                    "0-4 years", `Five-Year Age Groups`)
  )

who_weights <- tibble::tribble(
  ~age5,               ~w,
  "0-4 years",          0.0886,
  "5-9 years",          0.0869,
  "10-14 years",        0.0860,
  "15-19 years",        0.0847,
  "20-24 years",        0.0822,
  "25-29 years",        0.0793,
  "30-34 years",        0.0761,
  "35-39 years",        0.0715,
  "40-44 years",        0.0659,
  "45-49 years",        0.0604,
  "50-54 years",        0.0537,
  "55-59 years",        0.0455,
  "60-64 years",        0.0372,
  "65-69 years",        0.0296,
  "70-74 years",        0.0221,
  "75-79 years",        0.0152,
  "80-84 years",        0.0091,
  "85 years and over",  0.0064
)

df_age <- df %>%
  mutate(age5 = `Five-Year Age Groups`) %>%
  group_by(State, County, across(any_of("Year")), age5) %>%
  summarise(
    deaths = sum(Deaths, na.rm = TRUE),
    population = sum(Population, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  ungroup() %>%
  left_join(who_weights, by = "age5") %>%
  mutate(rate = (deaths / population) * 100000)

group_vars <- c("State", "County")
if ("Year" %in% names(df_age)) group_vars <- c(group_vars, "Year")

asmsr_county_us <- df_age %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(
    asmr_per_100k = sum(rate * w, na.rm = TRUE) / sum(w[!is.na(rate) & !is.na(w)], na.rm = TRUE),
    deaths_total  = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(State, County, across(any_of("Year")))


geo_lookup <- df_raw %>%
  transmute(
    STATE_NAME = State,
    COUNTY_NAME = County,
    STATEFP = `State Code`,
    COUNTYFP = `County Code`
  ) %>%
  mutate(COUNTY_NAME = clean_county_name(COUNTY_NAME)) %>%
  distinct(STATE_NAME, COUNTY_NAME, STATEFP, COUNTYFP) %>%
  add_geo_codes(state_col = STATEFP, county_col = COUNTYFP)

asmr_county_us <- asmsr_county_us %>%
  rename(STATE_NAME = State, COUNTY_NAME = County) %>%
  mutate(COUNTY_NAME = clean_county_name(COUNTY_NAME)) %>%
  left_join(geo_lookup, by = c("STATE_NAME","COUNTY_NAME")) %>%
  relocate(STATE_NAME, COUNTY_NAME, STATEFP, COUNTYFP, GEOID, COUNTRY)

head(asmr_county_us)

# write_csv(asmsr_county, "county_asmr_who.csv")

save(asmr_county_us, file = "county_asdmr_US.RData")

```

## Combined Border Region

### Crude Mortality Rate: USMX

load("USA_dm_mor_data.RData")

```{r USAMX, warning=FALSE, echo=FALSE}

load("Mexico_dm_mor_data.RData")
load("USA_dm_mor_data.RData")


Mexico_dm_mor_data <- Mexico_dm_mor_data %>%
  rename(
    COUNTY_NAME = `MUNICIPALITY_NAME`,
    STATEFP = `ENT_RESID`,
    COUNTYFP = `MUN_RESID`,
    Deaths= `total_deaths`,
    Population=`Total_Population`,
    `Crude Rate Lower 95% Confidence Interval`=`CDR_Lower_CI_Exact`,
    `Crude Rate Upper 95% Confidence Interval`= `CDR_Upper_CI_Exact`,
    `Crude Rate`= `Crude_Death_Rate`,
    `Crude Rate Standard Error`=`SE_rate`
  ) %>% 
   mutate(
    GEOID = paste0(STATEFP, COUNTYFP),
    COUNTRY = "Mexico"
  )

# Combine the datasets
USA_dm_mor_data <- USA_dm_mor_data %>%
  mutate(
    `Crude Rate` = as.numeric(`Crude Rate`)
  )


county_dm_mor_data_USMX <- bind_rows(USA_dm_mor_data, Mexico_dm_mor_data)

save(county_dm_mor_data_USMX, file = "county_dm_mor_data_USMX.RData")

```

### ASDMR: USMX

```{r}

load("county_asdmr_mx.RData")
load("county_asdmr_us.RData")

write_csv(asmr_county_mx, "county_asdmr_mx.csv")
write_csv(asmr_county_us, "county_asdmr_us.csv")

county_asdmr_mx <- asmr_county_mx %>%
  rename(
    COUNTY_NAME = `MUNICIPALITY_NAME`,
    STATEFP = `ENT_RESID`,
    COUNTYFP = `MUN_RESID`,
    asmr_per_100k= `ASMR`
  ) %>% 
   mutate(
    GEOID = paste0(STATEFP, COUNTYFP),
    COUNTRY = "Mexico"
  )

# just rename data to be consistent
county_asdmr_us <- asmr_county_us

# Combine the datasets
county_asdmr_USMX <- bind_rows(county_asdmr_us, county_asdmr_mx)

save(county_asdmr_USMX, file = "county_asdmr_USMX.RData")

```

## Map data setup

### Shape files

```{r warning=FALSE, message=FALSE}

# install.packages("rnaturalearth")

required_packages <- c(
  "sf", "maps", "mapdata", "usmap", "leaflet", "plotly",
  "viridis", "RColorBrewer", "scales", "tigris", "rmapshaper"
)

# Uncomment to install missing packages
install.packages(required_packages[!required_packages %in% installed.packages()])

library(sf)
library(ggplot2) 
library(dplyr)
library(viridis)
library(leaflet)
library(DT)
library(scales)
library(foreign)
library(rnaturalearth)
library(rnaturalearthhires)

library(maps)
library(mapdata)
library(usmap)
library(plotly)
library(RColorBrewer)
library(scales)
library(tigris)
library(rmapshaper)

# Set spatial options
options(tigris_use_cache = TRUE)
sf_use_s2(FALSE)  # Disable spherical geometry for simpler operations

# 1. Set path
file_path <- path_shapefiles

# 2. Load municipalities in Mexico and counties in USA

# Define US border states by their CVE_ENT codes
mx_border_states <- c(
  "02",  # Baja California
  "05",  # Coahuila
  "08",  # Chihuahua
  "19",  # Nuevo LeC3n
  "26",  # Sonora
  "28"   # Tamaulipas
)

us_border_states <- c(
  "04",  # Arizona
  "06",  # California  
  "35",  # New Mexico
  "48"   # Texas
)

mexico_mun <- st_read(file.path(file_path, "00mun.shp"), options = "ENCODING=LATIN1")
mexico_states <- st_read(file.path(file_path, "00ent.shp"), options = "ENCODING=LATIN1")

usa_counties <- st_read(file.path(file_path, "tl_2024_us_county.shp"))
usa_states <- st_read(file.path(file_path, "tl_2024_us_state.shp"))


# Rename columns in the mexico datasets
mexico_mun <- mexico_mun %>%
  rename(
    STATEFP = CVE_ENT,
    COUNTYFP = CVE_MUN,
    COUNTY_NAME = NOMGEO,
    GEOID = CVEGEO
  )

mexico_states <- mexico_states %>%
  rename(
    STATEFP = CVE_ENT,
    STATE_NAME = NOMGEO,
    GEOID = CVEGEO
  )

usa_states <- usa_states %>%
  rename(
    STATE_NAME = NAME
  )

usa_counties <- usa_counties %>%
  rename(
    COUNTY_NAME = NAME
  )

# st_geometry_type(usa_counties)
# st_geometry_type(mexico_mun)

# Standardize columns before binding
# names(mexico_mun) <- names(usa_counties)

# st_crs(usa_counties)
# st_crs(mexico_mun)

# Match mexico_mun to the CRS of usa_counties
usa_counties <- st_transform(usa_counties, st_crs(mexico_mun))
usa_states <- st_transform(usa_states, st_crs(mexico_states))


# Select only the common columns from each
usa_subset <- usa_counties %>%
  select(STATEFP, COUNTYFP, COUNTY_NAME, GEOID, geometry)%>%
  mutate(COUNTRY = "USA")

mexico_subset <- mexico_mun %>%
  select(STATEFP, COUNTYFP, COUNTY_NAME, GEOID, geometry)%>%
  mutate(COUNTRY = "Mexico")


usa_states_subset <- usa_states %>%
  select(STATEFP, STATE_NAME, GEOID, geometry) %>%
  mutate(COUNTRY = "USA")

mexico_states_subset <- mexico_states %>%
  select(STATEFP, STATE_NAME, GEOID, geometry) %>%
  mutate(COUNTRY = "Mexico")

# Append
usa_mexico_counties <- rbind(usa_subset, mexico_subset)
usa_mexico_states <- rbind(usa_states_subset, mexico_states_subset)

# Tag counties as part of US-Mexico border states
usa_mexico_counties <- usa_mexico_counties %>%
  mutate(
    is_us_border = case_when(
      COUNTRY == "USA" & STATEFP %in% us_border_states ~ "US-Mexico Border State",
      COUNTRY == "Mexico" & STATEFP %in% mx_border_states ~ "US-Mexico Border State",      TRUE ~ "Interior State"
    )
  )


excluded_statefps <- c(
  "02",  # Alaska
  "15",  # Hawaii
  "72",  # Puerto Rico
  "60",  # American Samoa
  "66",  # Guam
  "69",  # Northern Mariana Islands
  "78"   # U.S. Virgin Islands
)

# Filter counties
usa_mexico_counties_filtered <- usa_mexico_counties %>%
  filter(!(COUNTRY == "USA" & STATEFP %in% excluded_statefps))

map_data_counties_USMX <- usa_mexico_counties_filtered

saveRDS(map_data_counties_USMX, file.path(path_proc, "map_data_counties_USMX.rds"))

save(map_data_counties_USMX, file = "map_data_counties_USMX.RData")


# State-level data

# Prepare diabetes data for mapping (2021 prevalence)
# diabetes_map_data <- diabetes_data %>%
#   filter(
#     year == 2021,
#     metric_name == "Rate",
#     age_name == "All ages",
#     sex_name == "Both",
#     cause_name == "Diabetes mellitus",
#     measure_name == "Prevalence"
#   ) %>%
#   mutate(
#     # Standardize location names for matching
#     location_clean = case_when(
#       location_name == "Nuevo LeC3n" ~ "Nuevo Leon",
#       location_name == "Nuevo LeCB3n" ~ "Nuevo Leon",
#       TRUE ~ location_name
#     )
#   ) %>%
#   rename(
#     COUNTRY = country,
#     STATE_NAME = location_name
#   )
# 

usa_mexico_states_filtered <- usa_mexico_states %>%
  filter(!(COUNTRY == "USA" & STATEFP %in% excluded_statefps))

usa_mexico_states_filtered <- usa_mexico_states_filtered %>%
   mutate(
    STATE_NAME = ifelse(
      tolower(trimws(STATE_NAME)) == "coahuila de zaragoza",
      "Coahuila",
      STATE_NAME
    )
  )

saveRDS(usa_mexico_states_filtered, file.path(path_proc, "map_data_states_USMX.rds"))
# 
# 
# # join state-level diabetes data with shapefiles
# # Join using both STATE_NAME and COUNTRY
# 
# usa_mexico_states_filtered <- 
# 
# joined_diabetes_map_states_data <- left_join(
#   usa_mexico_states_filtered,
#   diabetes_map_data,
#   by = c("STATE_NAME", "COUNTRY")
# )
# 
# saveRDS(joined_diabetes_map_states_data, file.path(path_proc, "GBD_diabetes_2021_map_data_states_USMX.rds"))

```

### Crude Mortality Rate: USMX Map data

```{r}
# load datasets
load("county_dm_mor_data_USMX.RData")
map_data_counties_USMX <-readRDS(file.path(path_proc, "map_data_counties_USMX.rds"))

# join county-level diabetes data with shapefiles
# Join using both COUNTYFP, STATEFP, COUNTRY
joined_diabetes_map_county_data <- left_join(
  map_data_counties_USMX %>% select(-COUNTY_NAME, -GEOID),
  county_dm_mor_data_USMX,
  by = c("COUNTRY","STATEFP", "COUNTYFP")
)

# str(joined_diabetes_map_county_data)
# glimpse(joined_diabetes_map_county_data)


# Save as RData file (could save multiple objects) - Recommended for R
save(joined_diabetes_map_county_data, file = "joined_diabetes_map_county_data.RData")

```

### ASDMR: USMX Map data

```{r}
# load datasets
load("county_asdmr_USMX.RData")
load("map_data_counties_USMX.RData")

# county_asdmr_USMX %>%
#   count(COUNTRY, STATEFP, COUNTYFP) %>%
#   filter(n > 1)

# join county-level diabetes data with shapefiles
# Join using both COUNTYFP, STATEFP, COUNTRY
joined_asdmr_map_county_data <- left_join(
  map_data_counties_USMX %>% select(-COUNTY_NAME, -GEOID),
  county_asdmr_USMX,
  by = c("COUNTRY","STATEFP", "COUNTYFP")
)

# glimpse(joined_asdmr_map_county_data)


# Save as RData file (could save multiple objects) - Recommended for R
save(joined_asdmr_map_county_data, file = "joined_asdmr_map_county_data.RData")
```
